---
title: "Project 2: RNAseq"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
---

> **Note:** Sections marked **Instructions:** are scaffolding to guide you. Delete every
> instruction block once you've filled in the section, so the final document contains
> only your analysis and writing. Delete this section as well.
> If there is anything missing here or mistakes, please use the textbook guidelines as the
> ultimate source of truth.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
# Load the packages you use in this analysis here
library(tidyverse)
library(DESeq2)
library(fgsea)
```

```{r load_data}
# Read in your counts matrix, gene ID / name mapping, and sample metadata

```

# Phase 1: Blind analysis

## Experimental Design

> **Instructions (1 paragraph):** Using only the sample metadata you were given, describe
> the experiment:
>
> - How many groups are there, and how many replicates per group?
> - What covariates are provided (e.g. sex, batch, sequencing run)?
> - What comparison(s) can this design support? What kinds of effects might it be unable
>   to detect?
> - What potential confounders would you be concerned about, and can you account for them
>   with the information you have?
>
> *Delete this instruction block once you've completed this section.*



## Methods

> **Instructions (as long as needed):** Concisely describe which steps were taken in the
> analysis of the data, following our conventions for a methods section (see the methods
> guidelines presentation).
>
> - Specify the software versions used and the parameters for each step. If no parameters
>   were changed, state that default parameters were used.
> - Include any details necessary for someone else to replicate your analysis exactly.
> - You can exclude details that should not affect the results, such as exact file paths.
> - Cover both the Nextflow pipeline you developed and the analyses in this notebook.
>
> *Delete this instruction block once you've completed this section.*



## Read Quality Control

> **Instructions (1-2 paragraphs):** Use the MultiQC report or the individual FastQC and
> STAR logs to evaluate the quality of the reads. At minimum, address each of the following
> (even if there's no issue, state that there is no issue):
>
> - The range of the number of reads in all the samples
> - Any potential issues with the reads as flagged by FastQC
> - Any overrepresented sequences, or adapter contamination in the reads
> - The alignment rate of the reads to the reference genome
> - Multimapping rate
> - Whether you believe the experiment was of high quality and suitable for downstream
>   analysis. If not, what you would do to improve the quality of the reads?
>
> *Delete this instruction block once you've completed this section.*



## Filtering the Counts Matrix

> **Instructions (1-2 paragraphs):** Describe the filtering process you used on the counts
> matrix.
>
> - Include two plots showing the distribution of counts before and after filtering
> - State the number of genes present before and after filtering
> - Justify in a paragraph why you chose to filter in this way
>
> *Delete this instruction block once you've completed this section.*

```{r filter_counts}
# Filter the counts matrix

```

```{r counts_distribution_before}
# Plot the distribution of counts before filtering

```

```{r counts_distribution_after}
# Plot the distribution of counts after filtering

```

## Differential Expression Analysis

> **Instructions (3-4 paragraphs):**
>
> - Create a table of the top 10 differentially expressed genes and the statistics provided
>   by DESeq2, ranked by adjusted p-value
> - Choose a padj threshold and report the number of significant genes at this threshold,
>   separated into up- and downregulated
> - Use the thresholded results to perform a DAVID or Enrichr analysis on the significant
>   genes at your chosen padj threshold
>     - State the background gene set you used and justify it
>     - State whether you submitted up- and downregulated genes together or separately,
>       and why
> - Perform a GSEA analysis using
>   [fgsea](https://bioconductor.org/packages/release/bioc/html/fgsea.html) on your RNAseq
>   results using the C2 canonical pathways MSigDB dataset and log2 fold change as the
>   ranking metric
> - Choose a padj threshold for the fgsea analysis and create a plot of your choice that
>   displays the top most significantly enriched pathways
> - Comment briefly on the results of the DAVID or Enrichr analysis and the table you
>   created, and what it indicates about the biological processes that might differ
>   between groups
> - Comment briefly on the results of the fgsea analysis by stating what you believe are
>   the main significant pathways of interest and why
> - Compare the results from your DAVID/Enrichr analysis and your fgsea analysis and
>   comment on any similarities or differences you observe
>
> *Delete this instruction block once you've completed this section.*

```{r deseq2}
# Run DESeq2

```

```{r top10_table}
# Table of the top 10 DE genes ranked by padj

```

```{r significant_genes}
# Number of significant genes at your padj threshold, split into up- and downregulated

```

```{r enrichment_input}
# Prepare / write out gene lists and background for DAVID or Enrichr

```

## Sample Quality Control

> **Instructions (2 paragraphs):** Evaluate your samples before performing differential
> expression. These plots tell you whether your downstream results can be trusted.
>
> - Following the DESeq2 vignette, normalize the filtered counts using the rlog
>   transformation or the variance-stabilizing transformation
> - Perform PCA on the normalized counts matrix and overlay the sample information in a
>   biplot of PC1 vs. PC2
> - Create a heatmap or graphic of the sample-to-sample distances for the experiment
> - Do the samples cluster by experimental group? Are any samples outliers? Do any
>   covariates appear to explain the variation better than the experimental groups?
> - Based only on these plots, would you proceed to differential expression with all
>   samples as they are? If not, what would you change, and why?
>
> *Delete this instruction block once you've completed this section.*

```{r normalize}
# Normalize the filtered counts (rlog or VST)

```

```{r pca}
# PCA biplot of PC1 vs. PC2 with sample information overlaid

```

```{r sample_distances}
# Heatmap or graphic of sample-to-sample distances

```

## DAVID / Enrichr results



```{r fgsea}
# Run fgsea using C2 canonical pathways and log2 fold change as the ranking metric

```

```{r fgsea_plot}
# Plot the top significantly enriched pathways

```

## fgsea results



## Comparing DAVID/Enrichr and fgsea



## Interpretations

> **Instructions (2-3 paragraphs):** Using the number and direction of your DE genes, along
> with your pathway and enrichment results:
>
> - Speculate broadly on the major changes caused by the experimental condition
> - Which results most strongly support your speculation?
>
> *Delete this instruction block once you've completed this section.*



## Your Conclusions and Follow-up

> **Instructions:**
>
> - Propose two to three additional experiments you might do to validate your findings.
>   You do not have to do them; just propose them and anticipate what they would show you
>   if your findings held, and what result would convince you that your findings were
>   wrong.
> - Propose at least two follow-up experiments that would extend or expand on your
>   findings.
>
> *Delete this instruction block once you've completed this section.*

### Validation experiments



### Follow-up experiments



# Phase 2: Informed analysis

> **Note:** Do not revise your Phase 1 interpretation. Phase 2 is meant as a way to compare
> your results with the published work.

## Introduction

> **Instructions (1 paragraph):**
>
> - What is the biological background of the study?
> - Why was the study performed?
> - Why did the authors use the bioinformatic techniques they did?
>
> *Delete this instruction block once you've completed this section.*



## Replicate Figures 3C and 3F

> **Instructions (3-4 paragraphs):**
>
> - Create a volcano plot similar to the one seen in figure 3C
> - Use your DAVID/Enrichr or fgsea results and create a plot resembling figure 3F but with
>   your findings. You do not need to use the same pathways as they did.
> - Read their discussion of their results and address the numbered questions below.
>
> *Delete this instruction block once you've completed this section.*

```{r volcano_plot}
# Volcano plot similar to figure 3C

```

```{r figure_3f}
# Enrichment plot resembling figure 3F using your DAVID/Enrichr or fgsea results

```

### 1. Up- and downregulated genes

> **Instructions:** Compare how many significant genes are up- and downregulated in their
> findings and yours (using their significance threshold). List how many you find vs. how
> many they report.
>
> *Delete this instruction block once you've completed this section.*

```{r compare_counts}
# Number of up- and downregulated genes at the authors' significance threshold

```

|                | Your results | Reported in paper |
|----------------|--------------|-------------------|
| Upregulated    |              |                   |
| Downregulated  |              |                   |



### 2. Enrichment comparison

> **Instructions:** Compare their enrichment results with your DAVID/Enrichr and fgsea
> analyses. Comment on any differences you observe.
>
> *Delete this instruction block once you've completed this section.*



### 3. Sources of discrepancy

> **Instructions:** List the candidate sources of discrepancy between your results and
> theirs (e.g. genome and annotation versions, aligner, filtering, normalization,
> thresholds, multiple testing correction). Using the paper's methods as evidence, comment
> on at least two candidate sources you believe contribute the most to the differences you
> observe.
>
> - For each candidate source, propose a follow-up analysis or experiment that would reveal
>   whether it was the likely source of the differences. **You do not need to actually do
>   this analysis, just propose how.**
>
> *Delete this instruction block once you've completed this section.*



## Comparing Interpretations

> **Instructions (2-3 paragraphs):** Return to your first interpretations about the major
> biological pathways implicated in the results:
>
> - Did the authors emphasize any genes or pathways that were weak or absent in your
>   results? How strongly do you think their data support those claims?
> - Which of your findings agree with the authors'? Does reaching the same result through
>   a different pipeline make you more or less confident in it, and why?
>
> *Delete this instruction block once you've completed this section.*



## Conclusions

> **Instructions (1-2 paragraphs):**
>
> - Compare your first conclusions with those made by the paper. Did they propose any
>   follow-up experiments or perform any validation? What was different between what you
>   proposed and what they did or proposed in the text?
>
> *Delete this instruction block once you've completed this section.*



# Session Info

```{r session_info}
sessionInfo()
```
